---
title: "Weather Opens Opportunities in the North, Threatens Central and South"
subtitle: "Hay.Report for Jan 5, 2019"
output: ioslides_presentation
---

<!--
Welcome to the Hay Report for the week ending January 5, 2019. Hay dot Report analyzes the hay and forage market across the plains.
-->

## This Week

- No report this week from Corisca, SD, Colorado, or Kansas
- Mild weather in Northern Plains reducing usage and increasing exports
- Could a jump in costs be coming in the Central and Southern Plains?

<!--
There was no report this week from the Corsica hay auction, Colorado, or Kansas.
In this episode, weather is leading to opportunities for producers in the Northern Plains, but could a jump in prices be coming for the Central and Southern Plains?
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(dplyr)
library(tidyr)
library(ggplot2)

load('../data/report_data.rds')

CURRENT_DATE <- as.Date('2019-01-05')
CURRENT_WEEK <- lubridate::week(CURRENT_DATE)

state_lookup <- data.frame(state.name, state.abb)

region_lookup <- read.csv('~/hay.report/data/region_lookup.csv',
                          stringsAsFactors = FALSE)

report_data %>% 
  inner_join(state_lookup,
             by = c('state' = 'state.name')) %>% 
  left_join(region_lookup,
            by = c('state' = 'state',
                   'region' = 'region')) %>%
  filter(between(report_date, CURRENT_DATE - 31, CURRENT_DATE)) %>% 
  replace_na(list(size = '', shape = '', quality = '')) ->
  video_data

## Functions

week_ending <- function(DATE) {
  DATE + 7 - lubridate::wday(DATE)
}

graph_benchmark <- function(input_data, input_forage, 
                            input_shape, input_size, input_quality) {
  
  input_data %>%
  filter(forage == input_forage,
         shape == input_shape,
         size == input_size,
         quality == input_quality,
         !is.na(price_low)) %>%
  rowwise() %>% 
  mutate(price_max = max(c(price_low, price_high), na.rm = TRUE),
         report_week_ending = week_ending(report_date)) %>% 
  ungroup() %>% 
  ggplot(aes(x = report_week_ending, y = price_avg, 
             label = paste(region.abb, round(price_avg)),
            ymin = price_low, ymax = price_max
            )
         ) +
  geom_pointrange() +
  ggrepel::geom_text_repel(nudge_x = 1) +
  theme_bw() +
  labs(title = paste(input_forage, input_quality, input_size, input_shape),
       x = 'Week Ending',
        y = 'Price') 
}

benchmark_table <- . %>% 
    filter(!is.na(benchmark)) %>% 
  select(benchmark, report_year, report_week, price_low, price_high) %>% 
  gather(key = price_type, value = price, contains('price')) %>% 
  group_by(benchmark, report_year, report_week) %>% 
  summarize(benchmark_value = median(price, na.rm = TRUE)) %>% 
  group_by(benchmark) %>% 
  arrange(benchmark, report_year, report_week) %>% 
  summarize(current_value = last(benchmark_value) %>% 
              round(),
            one_week_change = ((last(benchmark_value) - first(benchmark_value))/first(benchmark_value)) %>% 
              magrittr::multiply_by(100) %>% 
              round() %>% 
              paste0('%')) %>% 
  knitr::kable(col.names = c('Commodity', 'Benchmark Price', '2 Wk Change'),
               align = 'lrr')
```

## Thank You

- Jack Carson, Oklahoma Dept of AG-USDA Market News, Oklahoma City, OK
- Heath Dewey, USDA-CO Dept of Ag Market News Service, Greeley, CO
- Lana Hutto, USDA Market News Service, Amarillo, TX
- John Kimbrell, USDA Market News Service, Billings, MT
- Ross Kotewa, USDA-SD Dept of Ag Market News Service, Sioux Falls, SD
- Kim Nettleton, Kansas Department of Agriculture, Manhattan, KS
- Thomas Walthers, USDA NE and USDA WY

<!--
Hay dot Report relies on the data and commentary of the USDA Agriculture Marketing Service. We thank them for their work serving farmers and ranchers across the country.
-->

## Benchmarks

- Key hay and forage types
- Multi-state & single state

<!--
Hay dot Report tracks several hay types to provide a broad view of the market. First is a look at alfalfa followed by grass hay. 
-->

## 

```{r alfalfa_benchmarks, eval = FALSE}
video_data %>% 
  filter(forage %in% c('Alfalfa', 'Dehydrated Alfalfa'),
         !is.na(price_low),
         between(report_date, CURRENT_DATE - 14, CURRENT_DATE)) %>% 
  mutate(benchmark = case_when(
    state == 'Kansas' & quality == 'Supreme' & shape == '' ~ 'Kansas Supreme Dairy',
    state == 'Texas' & quality == 'Premium/Supreme' & size == 'Large' & shape == 'Squares' ~ 'Texas Premium/Supreme',
    quality == 'Premium' & size == 'Large' & shape == 'Squares' ~ 'Multi-State Premium Large Squares',
    quality == 'Premium' & size == 'Small' & shape == 'Squares' ~ 'Multi-State Premium Small Squares',
    forage == 'Dehydrated Alfalfa' & quality == '17%' ~ 'Multi-State Dehydrated Pellets 17%',
    TRUE ~ NA_character_
  )) %>% 
  filter(!is.na(benchmark)) %>% 
  benchmark_table() ->
  alfalfa_table

save(alfalfa_table, file = paste0('../shared-data-products/alfalfa-table-', CURRENT_DATE, '.rda'))

alfalfa_table
```

```{r apls, fig.height=4.5}
video_data %>%
  graph_benchmark(input_forage = 'Alfalfa',
                  input_quality = 'Premium',
                  input_size = 'Large',
                  input_shape = 'Squares'
                  )

```

<!--
After the holidays, benchmark alfalfa commodities are steady. Premium Large Square median prices and ranges were unchanged from previous reports.
-->

##

```{r apls, fig.height=4.5}
video_data %>%
  #filter(state == 'Texas') %>% 
  graph_benchmark(input_forage = 'Alfalfa',
                  input_quality = 'Premium/Supreme',
                  input_size = 'Large',
                  input_shape = 'Squares'
                  )

```

<!--
Texas Premium Supreme Large Squares also remained steady. 
-->

##

```{r tpsls, fig.height=4.5}
video_data %>%
  graph_benchmark(input_forage = 'Alfalfa',
                  input_quality = 'Premium',
                  input_size = 'Small',
                  input_shape = 'Squares'
                  )

```

<!--
After ticking up for a week, median prices for Premium Small Squares ticked back down. The price spread between the Northern Plains and the High Plains of Texas is as large as the distance between the two regions. Even with the hundred dollar difference between regions and declining fuel prices, the eight hundred plus mile shipping cost may be too great for this commodity.
-->

##

```{r grass_benchmarks, eval = FALSE}
report_data %>% 
  filter(forage %in% c('Grass', 'Coastal Bermuda', 'Bluestem'),
         !is.na(price_low),
         between(report_date, CURRENT_DATE - 14, CURRENT_DATE)) %>% 
  mutate(benchmark = case_when(
    quality == 'Premium' & size == 'Small' & shape == 'Squares' ~ 'Multi-State Premium Small Squares',
    quality == 'Good' & size == 'Large' & shape == 'Rounds' ~ 'Multi-State Good Large Rounds',
    state == 'Texas' & forage == 'Coastal Bermuda' & size == 'Large' & shape == 'Rounds' & quality == 'Good/Premium' ~ 'Texas G/P Coastal Bermuda Large Rounds',
    forage == 'Bluestem' & size == 'Large' & shape == 'Squares' ~ 'Kansas Bluestem Large Squares',
    TRUE ~ NA_character_
  )) %>% 
  filter(!is.na(benchmark)) %>% 
  benchmark_table() ->
  grass_table

save(grass_table, file = paste0('../data-products/grass-table-', CURRENT_DATE, '.rda'))

grass_table
```

```{r cbpss, fig.height=4.5}
video_data %>%
  filter(state == 'Texas') %>% 
  graph_benchmark(input_forage = 'Coastal Bermuda',
                  input_quality = 'Good/Premium',
                  input_size = 'Large',
                  input_shape = 'Rounds'
                  )

```

<!--
Turning to grass, on December 14, Lana Hutto reported that: "Coastal Bermuda producers in East, North and South Texas have battled army worms, early frosts and rain in getting their 4th cutting therefore reports of short supply with many producers already sold out." As of now, prices are steady across the two regions.
-->

##

```{r cbpss, fig.height=4.5}
video_data %>%
  graph_benchmark(input_forage = 'Grass',
                  input_quality = 'Good',
                  input_size = 'Large',
                  input_shape = 'Rounds'
                  )

```

<!--
The price for Good Large Grass Rounds is largely steady, but prices at Corsica and in Eastern and Central Nebraska suggest a slight downward trend. 
-->

## Higher Prices on the Horizon?

AMS commentary suggests higher prices may be coming to the Central and Southern Plains

<!--
Despite prices being steady to a slight downward trend, there are concerns across the Central and Northern plains that recent weather may lead to higher prices.
-->

##

```{r cornstalks, fig.height=4.5}
video_data %>%
  filter(forage == 'Cornstalks') %>% 
  rowwise() %>% 
  mutate(price_max = max(c(price_low, price_high), na.rm = TRUE),
         report_week_ending = week_ending(report_date),
         Ground = ifelse(shape == 'Ground', 'Ground', 'Baled')) %>% 
  ungroup() %>% 
  ggplot(aes(x = report_week_ending, y = price_avg, 
             label = paste(region.abb, round(price_avg)),
            ymin = price_low, ymax = price_max
            )
         ) +
  geom_pointrange() +
  ggrepel::geom_text_repel(nudge_x = 1, aes(color = Ground)) +
  scale_color_manual(values = c('black', 'red')) +
  theme_bw() +
  labs(#title = paste(input_forage, input_quality, input_size, input_shape),
       x = 'Week Ending',
        y = 'Price') 

```

## Until Next Week

Visit www.hay.report for archives and resources.

<!--
Thank you for watching. Join Hay dot Report every Saturday for an update on the Plains hay and forage market. In the meantime, visit www.hay.report for archives. We love feedback. Please leave a comment with how Hay dot Report can continue to improve and serve you better.  
-->
